### 2014-10-01 paper/include/introduction.tex
stuff

DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption. This portion will continue to increase if nothing is done about it and a large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM. This refreshing has to be done as DRAM store data as a charge in a capacitor and given todays silicon, this charge leak over time. To increase DRAM performance and lower access latencies 3D stacked DRAM has been proposed in which the working temperature will be higher than in regular DRAM. This poses a problem as the rate at which the charge leaks increases with temperature and the DRAM will have to be refreshed more often.

@@

>>>>

DRAM is the most commonly used memory technique as main memory in computer systems today and represents a groving portion of system power consumption \cite{exascale}. This portion will continue to increase if nothing is done about it and a large part of the DRAM power consumption comes from refreshing all memory-cells in the DRAM. This refreshing has to be done as DRAM store data as a charge in a capacitor and given todays silicon, this charge leak over time. To increase DRAM performance and lower access latencies 3D stacked DRAM has been proposed in which the working temperature will be higher than in regular DRAM. This poses a problem as the rate at which the charge leaks increases with temperature and the DRAM will have to be refreshed more often.

@@

\begin{figure*}[t]
    \centering
    \subfigure[DRAM throughput loss]{
        \centering
        \includegraphics[width=0.45\textwidth]{dram_throughput}
		\label{fig:dram_throughput}
    }
    \subfigure[DRAM power consumption]{
        \centering
        \includegraphics[width=0.45\textwidth]{dram_power}
		\label{fig:dram_power}
    }%
	\caption{Projections on DRAM throughput loss and power consumption \cite{raidr}.}
	\label{fig:dram_data_proj}
\end{figure*}

### 2014-10-01 paper/include/rioparis.tex
stuff

RIO is a technique proposed by Seungjae Baek et al. \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them. This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them. They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average. As the implementaion only relies on changes in the kernel and there is no need to modify the hardware this technique would be easy to adopt in systems today. To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'. Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces. BLASDLBASDAHSUDGASJHDGJHASGDJHASGDJHAGSDJHGASJHDGAJSHDand it is temperature aware. Seungjae Baek et al. also propose PARIS discussed in \ref{sec:val} which can be combined with RIO for further improvments.

>>>>

RIO is a technique proposed by S. Baek et al. \cite{rioparis} in which they identify cells with low retention time (bad cells) and `deletes' them. This is done by changing the memory allocating modules in the OS to not allocate pages that have bad cells in them. They implement RIO in a Linux kernel which is run and tested on a real hardware system and is able to achieve 87.5\% reduction in refresh count with a performance increas of 4.5\% on average. It is shown that going for a refresh period longer than \(256ms\) only yields a small performance and power improvment and therefore they does not try to push it higher. Their implementaion almost only relies on changes in the kernel and there is no need to modify the hardware so this technique would be easy to adopt in systems today. To keep fragmentation of the memory low a maximum of 0.1\% of total memory pages gets `deleted'. Even though deletion of a larger amount of pages could decrease the amount of refreshes further it would cause trouble for the kernel which relies on being able to allocate large contiguous address spaces. As retentiontime varies hugely with temperature RIO continously measures the system temperature and compensate the refresh persiod for it. S. Baek et al. also propose PARIS which can be combined with RIO for further improvments. \todo[inline]{Write about paris}
